The role of advection versus diffusion in intraventricular transport

### 2025-03-06 paper/main.tex
Methods and end stuff.

\subsection{Quantities of interest}
We calculate the mean solute concentrations
\begin{equation}
    \cbar_i(t) = \frac{1}{|\Omega_i|}\int_{\Omega_i} c(\xx, t)\,\dx,
    \label{eq:c_mean_i}
\end{equation}
in the six regions of interest (ROIs) $\Omega_i$ ($i = 1, \dots, 6$)
as functions of time, where $|\Omega_i|$ is the volume of
the domain $\Omega_i$. The dynamics of ROI 1 follows directly
from~\Cref{eq:photoconversion_curve}. In addition,
we report time-to-threshold: the time $\hat{t}_i$ when
the mean concentration $\cbar_i$ first exceed a threshold
value $\hat{c}_i$. We use $\hat{c}_i=0.25$ for ROI 1--4 and
$\hat{c}_i=0.10$ for ROI 5--6.

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricles geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle.
To assess the assumption that inertial forces are negligible compared to
viscous forces in the CSF flow, such that the flow is governed by the Stokes equations,
we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r$ and $U_r$ as length and velocity scales characteristic of the flow.

When studying how cilia paralysis and ventricular morphology affect intraventricular
flow and transport, we calculate the magnitude of a tangential force $F$ 
exerted by the cilia on a part $\mathcal{S}\subset\Gc$ of the ventricular surface as:
\begin{equation}
    F = \int_{\mathcal{S}}\sqrt{\btau\cdot\btau}\,\mathrm ds.
    \label{eq:tangential_force_calculation}
\end{equation}

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}
The Stokes equations~\Cref{eq:stokes_eqs} are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization.
We set the timestep size $\Delta t = 0.02252$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM)
elements for the velocity.~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements,
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
To ensure stability of the scheme, continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity space is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly on each
element, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

\subsection{Numerical approximation of the advection-diffusion equation}
The advection-diffusion equation~\Cref{eq:adv_diff_strong} is discretized
with a first-order implicit Euler method in time, and a first-order,
symmetric interior penalty discontinuous Galerkin (DG)
method in space~\cite{Arnold1982AnElements}, chosen due to its favorable
conservation properties.
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.02252$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 25$ was chosen to ensure stability of the method.

\subsection{Solution strategy and software implementation}
Owing to the one-way coupling between the velocity field $\uu$
and the concentration $c$, the governing equations for the
CSF flow and the concentration can be solved sequentially when modeling solute transport.
For the cilia-only/no-cardiac flow scenario, the velocity quickly reaches steady-state since
the viscous forces dominate the inertial forces.
In this case, we first solve the steady-state Stokes equations~\Cref{eq:stokes_eqs}
for $\uu$ and $p$, and then use the velocity field $\uu$ as input when solving
the advection-diffusion equation~\Cref{eq:adv_diff_strong}.
For the other flow scenarios, the Stokes equations are solved for one
period of the pulsatile cardiac motion. The solution of this one period
is then used periodically as input for the advection-diffusion equation.

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with
DOLFINx~\cite{dolfinx2023preprint} with petsc4py~\cite{Dalcin2011ParallelPython}
as linear algebra backend.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we used an iterative
linear solver based on GMRES~\cite{Saad1986GMRES:Systems} with block Jacobi
preconditioning~\cite{Jacobi1845UeberGleichungen}. We discretize the time-dependent 
diffusion equation (without advection) and use the assembled linear system matrix
as the preconditioner matrix for the advection-diffusion problem.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}). The simulation code is openly
available~\cite{zenodo-link-to-all-code}\fixme{Add repo link.}.

@@

% Acknowledgements
\medskip
\textbf{Acknowledgments} \par 
Thanks are extended to J\o rgen Dokken for valuable model implementation discussions.
Halvor Herlyng acknowledges support from the national infrastructure for computational
science in Norway, Sigma2, via grant \#NN8049K.
This work was funded by the Research Council of Norway (NJY FRIPRO grants 314189 and 353348).

\medskip
\textbf{Data Availability Statement} \par
The data and implementations presented in this paper are licensed under \fixme{LICENSE}
and openly available at \fixme{REPO-LINK}.

\medskip
\textbf{Conflict of Interest Disclosure} \par
The authors declare no conflicts of interest.

\medskip
\textbf{Supporting Information} \par %Please delete the Suppporting Information statement
%if it is not applicable. Please supply Supporting Information in another file.
%Supporting information should not be provided in .tex format
Supporting Information is available from the Wiley Online Library or from the author.

% References
\medskip

>>>>

\subsection{Quantities of interest}
We calculate the mean solute concentrations
\begin{equation}
    \cbar_i(t) = \frac{1}{|\Omega_i|}\int_{\Omega_i} c(\xx, t)\,\dx,
    \label{eq:c_mean_i}
\end{equation}
in the six regions of interest (ROIs) $\Omega_i$ ($i = 1, \dots, 6$)
as functions of time, where $|\Omega_i|$ is the volume of
the domain $\Omega_i$. The dynamics of ROI 1 follows directly
from~\Cref{eq:photoconversion_curve}. In addition,
we report the time-to-threshold as the time $\hat{t}_i$ when
the mean concentration $\cbar_i$ first exceeds a threshold
value $\hat{c}_i$. We use $\hat{c}_i=0.25$ for ROI 1--4 and
$\hat{c}_i=0.10$ for ROI 5--6.

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricular geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle to yield $\overline{U} = \fixme{X}$ \fixme{unit}.
To assess the assumption that inertial forces are negligible compared to
viscous forces, we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r = \fixme{X}$ \fixme{unit} and $U_r = \fixme{X}$ \fixme{unit}
as length and velocity scales characteristic of the flow.

To study how cilia paralysis and ventricular morphology affect intraventricular
flow and transport, we calculate the magnitude of a tangential force $F$ 
exerted by the cilia on a part $\mathcal{S}\subset\Gc$ of the ventricular surface as:
\begin{equation}
    F = \int_{\mathcal{S}}\sqrt{\btau\cdot\btau}\,\mathrm ds.
    \label{eq:tangential_force_calculation}
\end{equation}

%-----DISCRETIZATION AND MODEL IMPLEMENTATION-----%
\subsection{Numerical approximation of the Stokes equations}
The Stokes equations (\Cref{eq:stokes_eqs}) are discretized with a finite element method
in space, while in time, we employ a first-order implicit Euler discretization.
We set the timestep size $\Delta t = 0.023$ s such that 20 timesteps represent one cardiac cycle.
For the spatial discretization, we use piecewise linear Brezzi-Douglas-Marini (BDM)
elements for the velocity.~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
The BDM elements have degrees of freedom associated with integral moments of the
velocity component normal to the facets of elements,
thus we enforce the impermeability condition $\uu\cdot\nn=0$ strongly.
We impose the boundary conditions for both the tangential traction $\bsigpar$ and
the normal pressure $\Tilde{p}(t)$ weakly.
For the pressure, we use piecewise constant (zeroth order discontinuous Galerkin (DG)) elements.
The resulting total number of degrees of freedom at each time step is 943 820.
The element combination $\mathrm{BDM}_1$--$\mathrm{DG}_{0}$ yields a non-conforming
discretization scheme for the Stokes equations~\cite{Stenberg1989SomeEquations}.
Continuity in tangential velocities on interior facets
is weakly enforced with a penalty parameter $\gamma=10$~\cite{Hong2016AEquations}.
Since the divergence of the velocity space is a subspace of the pressure space for this scheme,
the divergence-free condition of the velocity is satisfied exactly, ensuring mass conservation.~\cite{Boffi2008FiniteProblem}

\subsection{Numerical approximation of the advection-diffusion equation}
The advection-diffusion equation (\Cref{eq:adv_diff_strong}) is discretized
with a first-order implicit Euler method in time, and a first-order,
symmetric interior penalty DG
method in space~\cite{Arnold1982AnElements}, chosen due to its favorable
conservation properties.
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.023$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 25$ was chosen to ensure stability of the method.

\subsection{Solution strategy and software implementation}
Owing to the one-way coupling between the velocity field $\uu$
and the concentration $c$, the governing equations for the
CSF flow and the concentration can be solved sequentially.
For the cilia-driven/no-cardiac flow scenario, the velocity quickly reaches steady-state since
the viscous forces dominate the inertial forces.
In this case, we first solve the steady-state Stokes equations (\Cref{eq:stokes_eqs})
for $\uu$ and $p$, and then use the velocity field $\uu$ as input when solving
the advection-diffusion equation (\Cref{eq:adv_diff_strong}).
For the other flow scenarios, the Stokes equations are solved for one
period of the pulsatile cardiac motion. The solution of this one period
is then used periodically as input for the advection-diffusion equation.

The numerical solution schemes for the Stokes equations and the
advection-diffusion equation were implemented and solved numerically with
DOLFINx~\cite{dolfinx2023preprint} with petsc4py~\cite{Dalcin2011ParallelPython}
as linear algebra backend. The simulation code is openly
available\fixme{~\cite{zenodo-link-to-all-code}}.
A direct solver using MUMPS~\cite{Amestoy2011Mumps} was used to solve the
linear systems resulting from the discretized Stokes equations.
For the discretized advection-diffusion equation, we used an iterative
linear solver based on GMRES~\cite{Saad1986GMRES:Systems} with block Jacobi
preconditioning~\cite{Jacobi1845UeberGleichungen}. We discretize the time-dependent 
diffusion equation (without advection) and use the assembled linear system matrix
as the preconditioner matrix for the advection-diffusion problem.
Numerical experiments verified the accuracy of the model implementation
(\Cref{sec:appB_verification}).

@@

% Acknowledgements
\medskip
\noindent \textbf{Acknowledgments} \par
\noindent Thanks are extended to J\o rgen Dokken
for valuable model implementation discussions.  H.~H.~and
M.~E.~R. acknowledge support from the national infrastructure for
computational science in Norway, Sigma2, via grant
\#NN8049K. M.~E.~R.~acknowledges support from Stiftelsen Kristian
Gerhard Jebsen via the K.~G.~Jebsen Centre for Brain Fluid Research
and support from the Research Council of Norway (RCN) via grant
\#324239 (EMIx). N.~J.~Y.~acknowledges support from the Research
Council of Norway (grants \#314189 and \#353348).

\medskip
\noindent \textbf{Data Availability Statement} \par
\noindent The data and implementations presented in this paper are licensed under \fixme{LICENSE}
and openly available at \fixme{REPO-LINK}.

\medskip
\noindent \textbf{Conflict of Interest Disclosure} \par
\noindent The authors declare no conflicts of interest.

%\medskip
%\textbf{Supporting Information} \par %Please delete the Suppporting Information statement
%if it is not applicable. Please supply Supporting Information in another file.
%Supporting information should not be provided in .tex format
%Supporting Information is available from the Wiley Online Library or from the author.

% References
%\medskip
